---
title: "Group Project 1"
subtitle: "Biology 368/664 Bucknell University"
output: pdf_document
authors: Prof. Ken Field
date: 14 Sep 2024
---

```{r Load Libraries, include=FALSE}
# Load other packages here.
if (!require("datasauRus")) install.packages("datasauRus"); library(datasauRus)
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse)
```

This project will require you to develop a tutorial to teach Bucknell students how to use R for graphing and data analysis. 

## Introduction

Begin by introducing yourself to your group. 
Then discuss the biggest challenge that you have faced during the first three weeks of this course.
Determine if there are any common threads in these challenges and start to think about objectives for the tutorial.

## Target Audience

Discuss with your group the target audience for the tutorial. 
Examples could be a new student in Biology 364/664, a student in one of the new core Biology classes (201, 202, 203, or 204), a student in another 300-level Biology course (not 364), or a new student in one of the Bucknell research groups. 

Edit the README.md file in your group's repository to reflect your plan.

## Objectives

After deciding on the target audience for your tutorial determine 2 to 3 overall objectives for your tutorial (one per member of your group). 
These should be high-level objectives that are important skills for your target audience.
Check with Prof. Field to see if they are appropriate and then add them to your README.md file.

Identify at least 2 goals within each objective and add them to your README.md file.
These should be goals that someone who is working through the tutorial can self-asses.
For example, "to demostrate that you can test a hypothesis using a statistical model, the student should use a T test, linear model, or other test and interpret the p value appropriately."

## Grading

Each group will be expected to complete the following tasks to earn 85% of the points available for this assignment (21/25).

- Identify and obtain suitable dataset
- Use a Github repository and version control to collaborate on the project
  + Every member of the group should participate in editing the repo
- Spend 4-6 hours preparing, coding, and testing tutorial
  + Data exploration
  + Data visualization
  + Hypothesis testing
- Present tutorial in class
- Provide public archive suitable for sharing to students/faculty

Tutorials from previous classes can be viewed at our public github site: https://github.com/Bucknell-Biol364

Each group should use an *Acknowledgements* section to document the participation of each member and the collaboration within and between groups.

Additional credit will be awarded for providing assistance to other groups or for the development of a tutorial that goes beyond the minimal expectations listed above.
You will have the opportunity to provide feedback to another group after the initial deadline (like for Homework 02).

## Sample Dataset

One of the possible datasets to use for the tutorial can be found in the datasauRus package.

```{r}
datasaurus_dozen |> 
  group_by(dataset) |> 
  summarize(
      mean_x    = mean(x),
      mean_y    = mean(y),
      std_dev_x = sd(x),
      std_dev_y = sd(y),
      corr_x_y  = cor(x, y)
      )
```

Compare the means and standard deviations of the 13 different datasets.


Boxplots of either the x or the y value show that there are some differences, even though the means and standard deviations are identical.

```{r}
datasaurus_dozen |>
  ggplot(aes(x = x, colour = dataset)) +
    geom_boxplot() +
    theme_void() +
    theme(legend.position = "none") +
    facet_wrap(~dataset, ncol = 3)
```

```{r}
datasaurus_dozen |>
  ggplot(aes(x = y, colour = dataset))+
    geom_boxplot()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

But you have to visualize all of the data with a scatter plot to really see the patterns.

```{r fig.height=12, fig.width=9}
datasaurus_dozen |> 
  ggplot(aes(x = x, y = y, colour = dataset))+
    geom_point()+
    theme_void()+
    theme(legend.position = "none")+
    facet_wrap(~dataset, ncol = 3)
```

And did you notice the code in the {r} codechunk header that controlled the size of the output in the Rmd? Pretty neat trick!

And here are two versions of the data that you could use in your data visualization tutorial. 
To use them you would probably want to change the names of the datasets and also make x and y more meaningful. 
Then save them as a csv or tsv to be imported later for your tutorial. 

```{r}
datasaurus_long <- datasaurus_dozen
datasaurus_wide <- datasaurus_dozen_wide
head(datasaurus_long)
head(datasaurus_wide)
```

# Acknowledgements

DatasauRus package and description below: Stephanie Locke https://github.com/jumpingrivers/datasauRus

The datasauRus package wraps the awesome Datasaurus Dozen dataset, which contains 13 sets of x-y data. Each sub-dataset has five statistics that are (almost) the same in each case. (These are the mean of x, mean of y, standard deviation of x, standard deviation of y, and Pearson correlation between x and y). However, scatter plots reveal that each sub-dataset looks very different. The dataset is intended to be used to teach students that it is important to plot their own datasets, rather than relying only on statistics.

The Datasaurus was created by Alberto Cairo and described in the paper [Same Stats, Different Graphs: Generating Datasets with Varied Appearance and Identical Statistics through Simulated Annealing](https://www.autodeskresearch.com/publications/samestats) by Justin Matejka and George Fitzmaurice.

In the paper, Justin and George simulate a variety of datasets that the same summary statistics to the Datasaurus but have very different distributions. 

This package also makes these datasets available for use as an advanced [Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet), available in R as `anscombe`.

Data Exploration
Targeted audience: Students that have completed at least one elementary or basic course in R studio.


## Load in data
First we are going to load in our data using `read_csv()` which will turn a csv file into a dataframe.This code helps load the data for exploration and analyzing. We will also `view()` this data.

```{r, quiet = TRUE}
RxP <- read_csv("RxP.csv")
view(RxP)
```

## Manipulating data

This data set shows a lot of different variables but often we may not care about all of these variables. In order to `select()` data, we will use this command.
```{r}
RxP <- RxP |> 
  select(Hatch, Pred, Res, Mass.final, SVL.final) # Column names to be selected
```

When we view our RxP dataset now we will see that only the variables within `select()` are seen.

Lets take a look at our data using the `str()` command which returns the structure of our dataframe.
```{r}
str(RxP)
```

Data has different classes or types of values. Numeric refering to numbers, Characters words, and factors as levels. Currently our Hatch, Pred, and Res columns are considered characters but for graphing we will want to make sure they are `factors`. To do this we will use `mutate()`.
```{r}
RxP <- RxP |> 
  mutate(Hatch = as.factor(Hatch),
         Res = as.factor(Res),
         Pred = factor(Pred, levels = c("C", "NL", "L")))
```

`mutate` was able to take the data and change its type using `as.factor()`. Here we also used `factor()` and added `levels()` to the predation data. This will tell R that this data should be loaded in a specific order that being "C", then "NL", and then "L". This is less important now but will be helpful when it comes to graphing.

When we view the data again using `str()`, we will see these changes.
```{r}
str(RxP)
```

Now lets say we only want to view the largest of the tadpoles, we could use
`filter()` in order to do this
```{r}
BIG.RxP <- RxP |> 
  filter(Mass.final > 0.3) # Will remove any tadpoles that were had a final mass < 0.3
```

If you look in your environment in the top left of your screen you should now see significantly less observations (abbreviated as obs.) meaning we have successfully filtered data.

###############################################################
CAN SOMEONE DO SOMETHING ABOUT NORMALLY DISTRIBUTED DATA HERE
##############################################################

Because we can see that our data is not normally distributed, we might want to add a column that fixes this. We will be using `mutate()` again but in a different way to do this.
```{r}
RxP <- RxP |> 
  mutate(Log.mass = log10(Mass.final))
```

We have now created a new Log.mass column in or RxP dataframe. When we view this data with `simple.eda()` we will see that this data is a lot more normally distributed.

```{r}
simple.eda(RxP$Log.mass)
```

```{r}
str(RxP)
```


##Data Visualization
Data visualization is where we visually graph our data and gives more clarity to readers or viewers. So for this project we are going to graph three of the most common graphs in R studio, a bar graph, box plot, and linear graph. Now lets begin graphing our data.

The code below graph for bar graph. For bar graph, you have to use a factor and a numerical variable for better graph.This simply explains that hatching age is dependent on final mass.
```{r}
ggplot(RxP, aes(x = Hatch, y = Mass.final)) +
  geom_bar(stat = "identity") +
  labs(title = "Hatching age and Final Mass", x = "Hatching age", y = "Final Mass") +
  theme_cowplot()
```

Now it is your turn
start by calling the function ggplot. Replace the variables in the parenthesis to your data set
```{r}
ggplot(Data Set, aes(x = independent variable, y = dependent variable)) +
  geom_bar(stat = "identity") +
  labs(title = "what ever name you want to call your graph", x = "independent variable", y = "dependent variable") +
theme_cowplot()
```
##yayy you are done now.


The code below graph for box plot. For box plot, you also have to use a factor and a numerical variable for better graph. However, in this graph we compare three variable. hatching age and predation are dependent on final mass.
```{r}
ggplot(RxP, aes(x = Hatch, y = Mass.final)) +
  geom_boxplot() +
  labs(title = "Hatching age and Final Mass", x = "Hatching age", y = "Final Mass") +
  theme_cowplot()
```

Now let's practice!
start by calling the function ggplot.Replace the variables in the parenthesis to your data set.This time you are going to use a geom_boxplot function instead of geom_bar.
```{r}
ggplot(Data Set, aes(x = independent variable, y = dependent variable)) +
  geom_boxplot(stat = "identity") +
  labs(title = "what ever name you want to call your graph", x = "independent variable", y = "dependent variable") +
theme_cowplot()
```
##How are you feeling so far? We are about to something even cooler than our two graphs.

The code below graph for linear graph. For linear graph, you have to use two numerical variables for better graph. So we will use SVL.final and final mass. Linear graph helps show the relationship between variables. After our linear graph model, we then do our statistical test to explain our data.

```{r}
ggplot(RxP, aes(x = Hatch, y = Mass.final, fill = Pred)) +
  geom_boxplot() +
  labs(title = "Hatching age and Final Mass", x = "Hatching age", y = "Final Mass") +
  theme_cowplot()
```

Are you excited to compare more than two variables? Okay, great. The code you are about to use compares three variables all together and gives you a beautiful colorful graph. This time we will use fill to add our third varaible. Note this is a factor and not a numerical variable

```{r}
ggplot(Data Set, aes(x = independent variable, y = dependent variable, fill = Pred)) +
  geom_boxplot() +
  labs(title = "what ever name you want to call your graph", x = "independent variable", y = "dependent variable") +
theme_cowplot()
```


```{r linear graph}
ggplot(RxP, aes(x = SVL.final, y = Mass.final)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "SVL.final and Final Mass", x = "SVL.final", y = "Final Mass") +
  theme_cowplot()
```

Let's code for our last graph, which is linear graph.Call the ggplot function again and then use geom_point to construct a linear graph. However, for linear graph, you use numerical values only.alpha helps you to increase the visibility of the points along the line. You can play around that to see for yourself.
```{r}
ggplot(Data set, aes(x = numerical value 1, y = numerical value 2)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(method = "lm") +
  labs(title = "SVL.final and Final Mass", x = "SVL.final", y = "Final Mass") +
  theme_cowplot()
```
##Yayyy, Now we are done with data visualization. We can finally run our statistical test.

