---
title: "Group Project 1"
subtitle: "Biology 368/664 Bucknell University"
output: pdf_document
authors: Deborah Gonkpah, Jesse Gunn Cheu, Jack Strickland 
=======
date: 14 Sep 2024
---

```{r Load Libraries, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("UsingR")) install.packages("UsingR"); library(UsingR) # For simple.eda
if (!require("cowplot")) install.packages("cowplot"); library(cowplot)
if (!require("conflicted")) install.packages("conflicted"); library(conflicted) # For dealing with conflicts
if (!require("tidyverse")) install.packages("tidyverse"); library(tidyverse) # For everything
conflict_prefer_all("dplyr", quiet = TRUE)
```
##OBJECTIVE


## Load in data
First we are going to load in our data using `read_csv()` which will turn a csv file into a dataframe.This code helps load the data for exploration and analyzing. We will also `view()` this data.

=======
### Load in data

First we are going to load in our data using `read_csv()` which will turn a csv file into a dataframe. We will also `view()` this data.

```{r, quiet = TRUE}
RxP <- read_csv("RxP.csv") #loads in data as RxP
PRACTICE.RxP <- read_csv("RxP.csv") #loads in data as RxP
view(RxP)
```

Its important to note that the `<-` operator is how we assign variables rather than an `=` that many other computing languages use. 

**IMPORTANT VARIABLES IN THE DATA** 
`Predictor variables`
• Hatching age: Early (4 days post-oviposition) or Late (6 days post-oviposition)
• Predators: Control, Nonlethal (caged) dragonfly larvae, or Lethal (free-swimming)
dragonfly larvae
• Resources: Low (0.75 g) or High (1.5 g) food level, added every 5 days
• Block where the metamorph was reared
• Tank where the metamorph was reared
`Response variables`
• Age at hatching, both in terms of time since eggs were oviposited and time since
emergence began (defined as Day 1)
• Snout-vent (SVL) length at emergence
• Tail length at emergence
• SVL at completion of tail resorption
• Mass at completion of tail resorption
• Number of days needed for each metamorph to fully resorb the tail
*Touchon, Justin C., 'Basic Statistical Analyses using R', Applied Statistics with R: A Practical Guide for the Life Sciences (Oxford, 2021; online edn, Oxford Academic, 19 Aug. 2021), https://doi.org/10.1093/oso/9780198869979.003.0005, accessed 30 Sept. 2024.*


### Data Wrangling

This dataset shows a lot of different variables but more often than not we may not care about all of these variables. In order to `select()` data, we will use this command.
```{r}
RxP <- RxP |> 
  select(Hatch, Pred, Res, Mass.final, SVL.final) # Column names to be selected
```

When we view our RxP data set now, we will see that only the variables within `select()` are seen.

Lets take a look at our data using the `str()` command which returns the structure of our data frame. 

Note chr means character and num means numeric.
```{r}
str(RxP)
```


Data has different classes or types of values. Numeric refering to numbers, Characters words, and factors as levels. Currently our Hatch, Pred, and Res columns are considered characters but for graphing we will want to make sure they are `factors`. To do this we will use `mutate()`. 

`mutate` was able to take the data and change its type using `as.factor()`. Here we also used `factor()` and added `levels()` to the predation data. This will tell R that this data should be loaded in a specific order that being "C", then "NL", and then "L". This is less important now but will be helpful when it comes to graphing.

You may have also noticed an unfamiliar operator being used called a pipe ` |> `. Pipes are used in similar ways to parantheses and make code simpler to view. Here is an example of the same code using and not using pipes.
```{r}
str(select(RxP, Hatch, Mass.final))

RxP |> 
  select(Hatch, Mass.final) |> 
  str()
```

Now you `Select()` the `Tank`, `Block`, and `Tail.initial` columns.
```{r}
PRACTICE.RxP <- PRACTICE.RxP |> 
  select()

str(PRACTICE.RxP)
```

Good Job!

Data has different classes or types of values. Numeric refering to numbers, Characters words, and factors as levels. Currently our Hatch, Pred, and Res columns are considered characters but for graphing we will want to make sure they are `factors`. To do this we will use `mutate()`.

```{r}
RxP <- RxP |> 
  mutate(Hatch = as.factor(Hatch),
         Res = as.factor(Res),
         Pred = factor(Pred, levels = c("C", "NL", "L"))) 
```

When we view the data again using `str()`, we will see these changes, that is instead of just 26 variables, we see the 5 variables we are going to be using for our exploration. Also, this code this time tells you if the variable is a factor or numerical, which is very important for this data set. 

`mutate` was able to take the data and change its type using `as.factor()`. Here we also used `factor()` and added `levels()` to the predation data. This will tell R that this data should be loaded in a specific order that being "C", then "NL", and then "L". This is less important now but will be helpful when it comes to graphing.

Something that is also important to not in this above code chunk is the list of levels. Lists are created with `c()`. This data holds multiple items which can be incredibily useful as you develop your R skills. Here is a quick example of some more advanced code.
```{r}
for (column in c("Mass.final", "SVL.initial", "SVL.final", "Tail.initial")) {
  print(simple.eda(RxP[[column]]))
}
```


When we view the data again using `str()`, we will see these changes.

```{r}
str(RxP)
```

Practice using `factor()` and `mutate()` on a the `PRACTICE.RxP` dataset. Change the `Tank` and `Block` columns to factors.
```{r, quiet = TRUE}
PRACTICE.RxP <- PRACTICE.RxP |> 
  # YOUR CODE GOES HERE
```

Nice!

Now lets say we only want to view the largest of the tadpoles, we could use
`filter()` in order to do this
```{r}
BIG.RxP <- RxP |> 
  filter(Mass.final > 0.3) # Will remove any tadpoles that were had a final mass < 0.3
```

If you look in your environment in the top left of your screen you should now see significantly less observations (abbreviated as obs.) meaning we have successfully filtered data.


Shapiro-Wilk normality test. The normality test shows that final mass is not normally distributed.
=======
We can also do this with factors by determining if the factor meets certain specifications using comparison operators such as `==`.
```{r}
HIGH.RxP <- RxP |> 
  filter(Res == "Hi") # Will only include data that was in the high resource treatment groups
```

We can also do both of these filters at the same time.
```{r}
FILTER.RxP <- RxP |> 
  filter(Mass.final > 0.3,
         Res == "Hi")
```

Now its your turn, use `filter()` to select data that was in the `Control (C)` treatment group and had a `Final mass (final.mass)` less than 0.5.
```{r}
FILTER2.RxP <- # CODE GOES HERE
```

Wow!

These are the diagnostic plots below to check if things look okay or really out of whack. From these plots we can see that things do not look okay. The Q-Q Plot, boxplot, and histogram show a skewed data because the data may not be normally distributed. For example, the Q-Q Plot outliers at the far ends deviate from the diagonal line meaning there could be more extreme values than expected, hence, violating the model assumption. Lets do a quick check to see if some of our data is normally distributed using `simple.eda()`. `simple.eda()` is an easy visualization tool that allows us to view distributions of `numeric` data. Also the dollar sign is very important in the Shapiro-Wilk normality test. Always assign dollar sign to the variable you will be running a test on.
```{r}
simple.eda(RxP$Mass.final) # Show Mass Final
simple.eda(RxP$SVL.final)  # Show SVL Final
```

You try using `simple.eda()` with the `PRACTICE.RxP`.
```{r}
simple.eda() #Your code goes here
```

Amazing!



Because we can see that our data is not normally distributed, we might want to add a column that fixes this. We will be using `mutate()` again but in a log transformation way to do this.
```{r}
RxP <- RxP |> 
  mutate(Log.mass = log10(Mass.final))
```


We have now created a new Log.mass column in or RxP data frame. When we view this data with `simple.eda()` we will see that this data is a lot more normally distributed. log transformation helps the data to be more normally distributed. 

=======
We have now created a new Log.mass column in or RxP dataframe. When we view this data with `simple.eda()` we will see that this data is a lot more normally distributed.

```{r}
simple.eda(RxP$Log.mass)
```


##Data Visualization
Data visualization is where we visually graph our data and gives more clarity to readers or viewers. So for this project we are going to graph three of the most common graphs in R studio, a bar graph, box plot, and linear graph. Now lets begin graphing our data.

The code below graph for bar graph. For bar graph, you have to use a factor and a numerical variable for better graph.This simply explains that hatching age is dependent on final mass.
```{r}
ggplot(RxP, aes(x = Hatch, y = Mass.final)) +
  geom_bar(stat = "identity") +
  labs(title = "Hatching age and Final Mass", x = "Hatching age", y = "Final Mass") +
  theme_cowplot()
```


The code below graph for box plot. For box plot, you also have to use a factor and a numerical variable for better graph. However, in this graph we compare three variable. hatching age and predation are dependent on final mass.
```{r}
ggplot(RxP, aes(x = Hatch, y = Mass.final, fill = Pred)) +
  geom_boxplot() +
  labs(title = "Hatching age and Final Mass", x = "Hatching age", y = "Final Mass") +
  theme_cowplot()
```

The code below graph for linear graph. For linear graph, you have to use two numerical variables for better graph. So we will use SVL.final and final mass. Linear graph helps show the relationship between variables. After our linear graph model, we then do our statistical test to explain our data.

```{r linear graph}
ggplot(RxP, aes(x = SVL.final, y = Mass.final)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm") +
  labs(title = "SVL.final and Final Mass", x = "SVL.final", y = "Final Mass") +
  theme_cowplot()
```

=======
Lastly, you should take a look at `Tail.initial` data column using `simple.eda()` and determine if it is normally distributed. If you find that the data is not normally distributed, then you should create a new column using `mutate()` that finds the `log10()` of the data.
```{r}
#YOUR CODE GOES HERE
```

YIPPIE!

### VISUALIZATION

